---
title: "Probability Distributions in R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```
<style>

.tt {
  font-family:'Lucida Console', monospace; 
  text-align:center
}

h2 {
  color: darkgreen;
  font-weight: bold
}

h3 {
  color: darkolivegreen;
}

.topicsContainer {
  background-color: #f2f7f6;

}

.topic.current {
  background-color: #e6fcf8;
  border-color: #41ab98;
  border: solid 1px;
  font-weight: bold
}

.band {
  padding-left: 0%
}

.topicsFooter .resetButton {
  font-size: 80%;
}

.btn-primary, .btn-success, .btn-info {
    background-color: darkgreen;
    background-image: none;
    border: none;
}
    .btn-primary:hover, .btn-primary:focus, .btn-primary:active, .btn-primary:focus:active, .btn-success:hover, .btn-success:focus, .btn-success:active, .btn-success:focus:active, .btn-info:hover, .btn-info:focus, .btn-info:active, .btn-info:focus:active {
    background-color: darkolivegreen; }
  .btn-primary:focus:hover, .btn-primary:active:hover, .btn-primary:focus:active:hover, .btn-success:focus:hover, .btn-success:active:hover, .btn-success:focus:active:hover, .btn-info:focus:hover, .btn-info:active:hover, .btn-info:focus:active:hover {
    background-color: darkolivegreen; }
  .btn-primary.disabled, .btn-primary:disabled, .btn-primary.disabled:hover, .btn-primary:disabled:hover, .btn-success.disabled, .btn-success:disabled, .btn-success.disabled:hover, .btn-success:disabled:hover, .btn-info.disabled, .btn-info:disabled, .btn-info.disabled:hover, .btn-info:disabled:hover {
    background-color: darkolivegreen; }
    
.boxed{
 border: 3px solid green ;
 padding: 5px;
}


</style>

## Probability Distributions in R

### Introduction

In this tutorial we learn about various common probability distributions used in R that have pre-made functions built in to R for solving various problems.

For each distribution we will:

* Review the distribution's pdf, shape, and parameters

* Use R to calculate the probability of falling in a range of values

* Use R to calculate quantiles for the distribution

* Use R to find the value of the PDF at a particular $x$

* Generate random values from the distribution


## The Uniform Distribution

### The Parameters and PDF of a Continuous Uniform Distirbution

In a (continuous) uniform distribution, all points in the sample space are equally likely to occur. This means that our PDF is a flat line and the area under the PDF is a rectangle. To completely identify a uniform distribution, we need to know the range of values our variable can take on. For a random variable $X$ that follows a uniform distribution that takes on values between $a$ and $b$ we write:

$$X\sim U(a, b)$$

and express the PDF as:

\begin{equation*}
f(x)=\begin{cases}
          \dfrac{1}{b-1} \quad &\text{if} \, x \in \left[a,b\right] \\
          0 \quad &\text{otherwise}\\
     \end{cases}
\end{equation*}

The graph of the PDF is:

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
dat <- data.frame(x1 = 2, x2 = 10, y1 = 1/8, y2 = 1/8)
ggplot(dat) +
  geom_segment(aes(x=x1, y = y1, xend = x2, yend = y2)) +
  ylim(0, 0.150) +
  xlab("x") +
  ylab("f(x)") +
  scale_x_continuous(breaks= c(2, 10),
              labels= c("a", "b")) +
  ggtitle("PDF of X~U(a,b)") +
  theme_classic() +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14))
```

### Area to the Left

To find the probability of being less than a point $q$, i.e.

$$ P(X<q) = \int_{-\infty}^qf(t)dt,$$
we can use the 

<p class = "tt">punif(q, min = a, max = b)</p>

function. If no values are specified for $a$ and $b$, R will assume you want the interval $(0,1)$ by default.

<div class = "example">
**Example** Suppose $X$ follows a uniform distribution from 7 to 23. What is the probability that $X$ is less then 19?

</div>

<div class = "solution">
**Solution** $P(X<19)=$

```{r, echo = TRUE}
punif(19, 7, 23)
```

Visually, this looks like:

```{r, warning = FALSE, message = FALSE}
f <- function(x){
  if(7<= x & x <= 23) y <- 1/16
  else y <- 0
  return(y)
}

x = seq(from=7,to=23,by=.005)
y = sapply(X=x,FUN=f,simplify=TRUE,USE.NAMES=FALSE)
df = data.frame(X=x,Y=y)

ggplot(data=df, mapping=aes(x=X, y=Y)) +
    geom_line()+
    geom_area(mapping=aes(x=ifelse(x>7 & x<19, x, 0)), fill="lightblue", alpha=1, color = "black") +
    ylim(0, 0.07) +
    xlim(5, 25) +
   xlab("x") +
    ylab("f(x)") +
    ggtitle("P(X<19) for X~U(7,23)") +
   theme_classic() +
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=14))

# dat <- data.frame(x1 = 7, x2 = 23, y1 = 1/16, y2 = 1/16)
# ggplot(dat) +
#   geom_segment(aes(x=x1, y = y1, xend = x2, yend = y2)) +
#   ylim(0, 0.07) +
#   xlab("x") +
#   ylab("f(x)") +
#   ggtitle("PDF of X~U(7,23)") +
#   theme_classic()
```
</div>

## The Normal Distribution


## The Binomial Distribution